---
title: "Getting Started with rSuStaIn"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with rSuStaIn}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Introduction

The `rSuStaIn` package provides an R interface to the 
[pySuStaIn](https://github.com/ucl-pond/pySuStaIn) Python package for 
Subtype and Stage Inference (SuStaIn). SuStaIn is an unsupervised learning 
algorithm that identifies disease subtypes and stages from cross-sectional data.

# Installation

First, install the R package from GitHub:

```{r}
# install.packages("devtools")
devtools::install_github("ucdavis/rSuStaIn")
```

Then, install the pySuStaIn Python package:

```{r}
library(rSuStaIn)
install_pySuStaIn()
```

This will create a Python virtual environment called "r-pySuStaIn" and install
pySuStaIn and its dependencies.

# Basic Usage

## Preparing Your Data

SuStaIn requires:

1. **Patient data**: A data frame with biomarker measurements for each subject
2. **Biomarker levels**: A list defining the possible levels for each biomarker
3. **Probability of correct classification**: Estimates of measurement accuracy

```{r}
# Example data structure
patient_data <- data.frame(
  ID = paste0("Patient_", 1:100),
  biomarker1 = sample(c("Normal", "Mild", "Severe"), 100, replace = TRUE),
  biomarker2 = sample(c("Normal", "Mild", "Severe"), 100, replace = TRUE),
  biomarker3 = sample(c("Normal", "Mild", "Severe"), 100, replace = TRUE)
)

# Define biomarker levels
biomarker_levels <- list(
  biomarker1 = c("Normal", "Mild", "Severe"),
  biomarker2 = c("Normal", "Mild", "Severe"),
  biomarker3 = c("Normal", "Mild", "Severe")
)
```

## Running the Ordinal SuStaIn Algorithm

```{r}
# Run SuStaIn
results <- run_OSA(
  patient_data = patient_data,
  biomarker_levels = biomarker_levels,
  SuStaInLabels = names(biomarker_levels),
  N_startpoints = 25,        # Number of startpoints for optimization
  N_S_max = 3,               # Maximum number of subtypes
  N_iterations_MCMC = 1e5,   # MCMC iterations
  output_folder = "output",
  dataset_name = "my_analysis"
)
```

## Interpreting Results

The results object contains:

- `samples_sequence`: Sampled biomarker event sequences for each subtype
- `samples_f`: Sampled subtype frequencies
- `ml_subtype`: Maximum likelihood subtype assignment for each subject
- `prob_ml_subtype`: Probability of maximum likelihood subtype
- `ml_stage`: Maximum likelihood stage assignment for each subject
- `prob_ml_stage`: Probability of maximum likelihood stage

```{r}
# View subtype assignments
head(results$ml_subtype)

# View stage assignments
head(results$ml_stage)
```

## Working with Pickle Files

If you've already run SuStaIn in Python and have pickle files, you can extract
the results:

```{r}
results <- extract_results_from_pickle(
  n_s = 3,                    # Number of subtypes
  dataset_name = "my_data",
  output_folder = "output"
)
```

# Advanced Usage

## Cross-Validation

You can perform cross-validation to assess model fit:

```{r}
results <- run_OSA(
  patient_data = patient_data,
  biomarker_levels = biomarker_levels,
  SuStaInLabels = names(biomarker_levels),
  N_startpoints = 25,
  N_S_max = 3,
  N_CV_folds = 10,           # Enable 10-fold cross-validation
  output_folder = "output",
  dataset_name = "my_analysis"
)

# Access CV results
cv_results <- attr(results, "CV")
```

## Customizing Probability of Correct Classification

You can specify custom probabilities for correct classification:

```{r}
prob_correct <- compute_prob_correct(
  biomarker_levels = biomarker_levels,
  # Customize parameters here
)

results <- run_OSA(
  patient_data = patient_data,
  prob_correct = prob_correct,
  biomarker_levels = biomarker_levels,
  # ... other parameters
)
```

# References

Young, A. L., Vogel, J. W., Aksman, L. M., Wijeratne, P. A., Eshaghi, A., 
Oxtoby, N. P., ... & Alzheimer's Disease Neuroimaging Initiative. (2021). 
Ordinal SuStaIn: Subtype and Stage Inference for Clinical Scores, Visual Ratings, 
and Other Ordinal Data. Frontiers in Artificial Intelligence, 4, 613261.
